An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-07-25 latex_files/ch1/1_1_motivation.tex
fixed warnings

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism \cite{Gutmann2016}. The key characteristic
of these models is that although we can sample (simulate) data points,
we cannot evaluate the likelihood of a specific set of observations
$\data$. Formally, a simulator-based model is described as a
parameterised family of probability density functions
$\{ p_{\yb|\thetab}(\yb) \}_{\thetab}$, whose closed-form is either
unknown or intractable to evaluate. Whereas evaluating
$p_{\yb|\thetab}(\yb)$ is intractable, sampling is
feasible. Practically, a simulator can be understood as a black-box
machine $M_r$\footnote{The subscript $r$ in $M_r$ indicates the
  \textit{random} simulator. In the next chapters we will introduce
  $M_d$ indicating the \textit{deterministic} simulator.} that
given a set of parameters $\thetab$, produces samples $\yb$ in a
stochastic manner i.e.\ $M_r(\thetab) \rightarrow \yb$.

>>>>

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism \cite{Gutmann2016}. The key characteristic
of these models is that although it is feasible to sample (simulate) data points, it is impossible to evaluate the likelihood. Formally, a simulator-based model is a
parameterised family of probability density functions
$\{ p_{\yb|\thetab}(\yb) \}_{\thetab}$, whose closed-form is either
unknown or intractable to evaluate. Practically, a simulator can be understood as a black-box
machine $M_r$\footnote{The subscript $r$ in $M_r$ indicates the
  \textit{random} simulator. In the next chapters we will introduce
  $M_d$ indicating the \textit{deterministic} simulator.} that
given a set of parameters $\thetab$, produces samples $\yb$ in a
stochastic manner i.e.\ $M_r(\thetab) \rightarrow \yb$.

### 2021-07-25 latex_files/ch3/3_2_1_general_design.tex
fixed warnings

The overview of our implementation is illustrated in
figure~\ref{graph:romc_overview}; one may interpret
figure~\ref{graph:romc_overview} as a depiction of the main class and
the ellipses are the main routines of the class. Following
\proglang{Python}'s naming principles, the methods starting with an
underscore (green ellipses) represent internal (private) functions,
whereas the rest (blue ellipses) are the public methods that the user
interacts with.

@@

In figure~\ref{graph:romc_overview}, it can be easily observed that the
implementation follows consistently the algorithmic description
described above, which follows the partition into training, inference
and evaluation. The training part includes all the steps until the
computation of the proposal regions i.e. sampling the nuisance
variables, defining the optimization problems, solving them,
constructing the regions and fitting local surrogate models. The
inference part comprises of evaluating the unnormalized posterior (and
the normalized when is possible), sampling and computing an
expectation. From an algorithmic point of view, the training and the
inference parts are \emph{isolated}. We also provide some utilities
for inspecting the training process, such as plotting the histogram of
the final distances or visualizing the constructed bounding
boxes. Finally, in the evaluation part, we provide two methods for
evaluating the inference; (a) computing the Effective Sample Size
(ESS) of the samples and (b) measuring the divergence between the
approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method
  (i.e.\ ABC Rejection Sampling), and we would like to measure the
  discrepancy between the two approximations.}

>>>>

@@

The overview of our implementation is illustrated in
figure~\ref{fig:romc_overview}; one may interpret
figure~\ref{fig:romc_overview} as a depiction of the main class and
the ellipses are the main routines of the class. Following
\proglang{Python}'s naming principles, the methods starting with an
underscore (green ellipses) represent internal (private) functions,
whereas the rest (blue ellipses) are the public methods that the user
interacts with.

In figure~\ref{fig:romc_overview}, it can be easily observed that the
implementation follows consistently the algorithmic description
described above, which follows the partition into training, inference
and evaluation. The training part includes all the steps until the
computation of the proposal regions i.e. sampling the nuisance
variables, defining the optimization problems, solving them,
constructing the regions and fitting local surrogate models. The
inference part comprises of evaluating the unnormalized posterior (and
the normalized when is possible), sampling and computing an
expectation. From an algorithmic point of view, the training and the
inference parts are \emph{isolated}. We also provide some utilities
for inspecting the training process, such as plotting the histogram of
the final distances or visualizing the constructed bounding
boxes. Finally, in the evaluation part, we provide two methods for
evaluating the inference; (a) computing the Effective Sample Size
(ESS) of the samples and (b) measuring the divergence between the
approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method
  (i.e.\ ABC Rejection Sampling), and we would like to measure the
  discrepancy between the two approximations.}

### 2021-07-25 latex_files/ch3/3_2_3_inference.tex
fixed warnings

In the following code snippet, we perform weighted sampling from the
ROMC approximate posterior. Afterwards, we used some of ELFI's
built-in tools to get a summary of the obtained samples. In figure
\ref{fig:example_sampling}, we observe the histogram of the weighted
samples and the acceptance region of the first deterministic function
(as before) alongside the samples obtained from it. Finally, in the
code snippet, we demonstrate how to use the \code{compute_expectation}
function; in the current example, we define \code{h} to compute
firstly the empirical mean and afterwards the empirical variance. In
both cases, the empirical result is close to the ground truth
$\mu = 0$ and $\sigma^2 = 1$. The
\code{romc.eval_unnorm_posterior(theta)} evaluates the posterior at
point $\theta$ using expression \eqref{eq:aprox_posterior}. The
\code{romc.eval_posterior(theta)} approximates the partition function
$Z = \int_{\thetab} p_{d,\epsilon}(\thetab|\data) d\thetab$ using the
Riemann approximation as explained above. In our simple example, this
utility can provide a nice plot of the approximate posterior as
illustrated in figure~\ref{fig:approx_posterior}. We observe that the
approximation is quite close to the ground-truth posterior.

>>>>

In the following code snippet, we perform weighted sampling from the
ROMC approximate posterior. Afterwards, we use some of ELFI's
built-in tools to get a summary of the obtained samples. In figure
\ref{fig:running_example_romc_inference}, we observe the histogram of the weighted
samples and the acceptance region of the first deterministic function
(as before) alongside the samples obtained from it. Finally, in the
code snippet, we demonstrate how to use the \code{compute_expectation}
function; in the current example, we define \code{h} to compute
firstly the empirical mean and afterwards the empirical variance. In
both cases, the empirical result is close to the ground truth
$\mu = 0$ and $\sigma^2 = 1$. The
\code{romc.eval_unnorm_posterior(theta)} evaluates the posterior at
point $\theta$ using expression \eqref{eq:approx_posterior_omc}. The
\code{romc.eval_posterior(theta)} approximates the partition function
$Z = \int_{\thetab} p_{d,\epsilon}(\thetab|\data) d\thetab$ using the
Riemann approximation as explained above. In our simple example, this
utility can provide a nice plot of the approximate posterior as
illustrated in figure~\ref{fig:running_example_romc_inference}. We observe that the
approximation is quite close to the ground-truth posterior.

### 2021-07-25 latex_files/ch4/4_2_ma2.tex
fixed warnings

\begin{figure}[h]
    \begin{center}
      % \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
      \input{./latex_files/images/chapter4/ma2_region_3.tex}
      \input{./latex_files/images/chapter4/ma2_region_3_bo.tex}
      % \includegraphics[width=0.35\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesia Optimisation.}
  \label{fig:ma2_5}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
    \input{./latex_files/images/chapter4/mae2_hist_t1_rejection.tex}
    \input{./latex_files/images/chapter4/mae2_hist_t2_rejection.tex}\\
    \input{./latex_files/images/chapter4/mae2_hist_t1_romc.tex}
    \input{./latex_files/images/chapter4/mae2_hist_t2_romc.tex}\\
    \input{./latex_files/images/chapter4/mae2_hist_t1_romc_bo.tex}
    \input{./latex_files/images/chapter4/mae2_hist_t2_romc_bo.tex}\\
    \end{center}
    \caption[MA2 example, evaluation of the marginal
    distributions.]{Histogram of the marginal posterior distributions
      using three different inference approaches; (a) in the first
      row, the samples are obtained using Rejection ABC sampling (b)
      in the second row, using ROMC with a gradient-based optimiser
      and (c) in the third row, using ROMC with Bayesian optimization
      approach. The vertical (red) line represents the samples mean
      $\mu$ and the horizontal (black) the standard deviation
      $\sigma$.}
  \label{fig:ma2_3}
\end{figure}

>>>>

\begin{figure}[ht]
    \begin{center}
        % \input{./latex_files/images/chapter4/ma2_region_3.tex}
        % \input{./latex_files/images/chapter4/ma2_region_3_bo.tex}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesian Optimisation.}
  \label{fig:ma2_5}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \input{./latex_files/images/chapter4/mae2_hist_t1_rejection.tex}
    \input{./latex_files/images/chapter4/mae2_hist_t2_rejection.tex}\\
    \input{./latex_files/images/chapter4/mae2_hist_t1_romc.tex}
    \input{./latex_files/images/chapter4/mae2_hist_t2_romc.tex}\\
    \input{./latex_files/images/chapter4/mae2_hist_t1_romc_bo.tex}
    \input{./latex_files/images/chapter4/mae2_hist_t2_romc_bo.tex}\\
    \end{center}
    \caption[MA2 example, evaluation of the marginal
    distributions.]{Histogram of the marginal posterior distributions
      using three different inference approaches; (a) in the first
      row, the samples are obtained using Rejection ABC sampling (b)
      in the second row, using ROMC with a gradient-based optimiser
      and (c) in the third row, using ROMC with Bayesian optimization
      approach. The vertical (red) line represents the samples mean
      $\mu$ and the horizontal (black) the standard deviation
      $\sigma$.}
  \label{fig:ma2_3}
\end{figure}
